Persistent somatic symptoms are key to individual illness perception at one year after COVID-19 in a cross-sectional prospective cohort study

### 2023-02-09 paper/markdown/manuscript.Rmd
Second revision
Minor Reviewer Issues, duplicated references removed

A subset of coronavirus disease 2019 (COVID-19) patients is affected by protracted somatic symptoms, cardiopulmonary pathology and mental health disorders [@Sonnweber2020; @Sahanic2021; @Luger2022; @Sonnweber2022; @Hufner2022; @Al-Aly2021; @Davis2021; @Soriano2022; @Evans2021a; @Sudre2021; @Chen2022]. 
Persistent COVID-19-related symptoms have been initially described by patient initiatives [@McCorkell2021] and subsequently recognized as 'post-COVID-19 condition' by the clinical community [@Sonnweber2020; @Sahanic2021; @Davis2021; @Soriano2022; @Al-Aly2021; @Sudre2021]. 
<span custom-style = 'revision'>Most patients recover completely from acute COVID-19 symptoms within four weeks. 
The clearance rate is particularly fast for non-specific infection complaints [@Sudre2021; @Sahanic2021; @Rass2022]. 
In a recent metaanalysis of 1.7 million patients, the pooled prevalence estimate of post-COVID-19 condition, i.e. symptoms present for 3 months or longer, was 43% [@Soriano2022; @Chen2022]. 
Yet, in a rigorously controlled study, approximately 12.7% of SARS-CoV-2-positive individuals suffered from persistent somatic symptoms directly associated with COVID-19 [@Ballering2022].
Fatigue, concentration and memory problems, sleep disorders, physical performance deficits, smell/taste disorders and hair loss, dyspnea and pain are the most common, frequently overlapping and relapsing manifestations [@Chen2022; @Sonnweber2022; @Sahanic2021; @Davis2021; @Rass2022; @Evans2021a; @Huang2021; @Ballering2022].</span>

@@

<span custom-style = 'revision'>We aimed to describe IP and IP components, identify demographic and clinical variables affecting IP and explore IP patterns in convalescents suffering from persistent somatic symptoms (PSS), lung or heart abnormalities one year after COVID-19.
To this end, we gauged IP in a long-term observatory cross-sectional COVID-19 collective [@Sonnweber2020; @Sahanic2021; @Luger2022; @Sonnweber2022] with the Brief Illness Perception Questionnaire (BIPQ) [@Broadbent2006] and investigated components, influential factors and patterns of IP by factor analysis, regularized multi-parameter modeling and unsupervised PAM clustering.</span>

@@

<span custom-style = 'revision'>In total, 190 individuals were screened for participation. 
The screened subset size was limited by the number of COVID-19 cases diagnosed at the study centers, their follow-up availability (residency in the study region) and available resources for the follow-up assessments. 
No statistical sample size estimation was performed, as the kinetic of COVID-19 recovery was unknown at the pandemic onset.
Out of 145 enrolled individuals, `r nrow(cov_data$raw_data)` completed the one-year follow-up.</span>
A subset of participants (n = `r nrow(cov_data$clear_data)`) was analyzed who displayed (1) PSS or (2) any abnormality in chest computed tomography (CT) or (3) any lung function testing (LFT) deficits or (4) any heart abnormality in trans-thoracic echocardiography (TTE) at the one-year follow-up <span custom-style = 'revision'>and had a complete set of BIPQ, demographic and clinical analysis variables (__Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1 - S2__).</span>
<span custom-style = 'revision'>We decided to restrict the analysis to incomplete COVID-19 recovery, because BIPQ was initially designed for and validated in somatic conditions [@Broadbent2006].</span>

@@

Physical performance was rated with the Eastern Cooperative Oncology Group scale (ECOG). 
Dyspnea was scored with the Modified Medical British Research Council scale (mMRC).
Fatigue was rated with likert and bimodal Chalder's Fatigue Scales (CFS) [@Chalder1993; @Morriss1998]. 
Exertional capacity was assessed by six-minute walking distance (SMWD) [@Crapo2012].

The following, <span custom-style = 'revision'>PSS,</span> were analyzed at the one-year follow-up: 
reduced physical performance (ECOG $\geq$ 1), 
dyspnea (mMRC $\geq$ 1), 
self-reported cough (yes/no item), 
self-reported sleep problems (yes/no), 
self-reported night sweating (yes/no), 
self-reported hyposmia or anosmia (yes/no), 
self-reported dermatological symptoms (yes/no), 
self-reported gastrointestinal symptoms (yes/no), 
self-reported hair loss (yes/no), 
significant fatigue (bimodal CFS $\geq$ 4). 
<span custom-style = 'revision'>Records of reduced physical performance, dyspnea, cough, sleep problems, night sweating, hyposmia/anosmia and gastrointestinal symptoms were available for acute COVID-19 and all follow-ups [@Sonnweber2022].</span>

LFT abnormality was defined as at least one parameter < 80% (forced vital capacity [FVC], forced expiratory volume in 1 second [FEV1], total lung capacity, diffusion lung capacity for carbon monoxide) or < 70% (FEV1:FVC ratio) of the reference value [@Sonnweber2022]. 
CT images were evaluated with the Fleischner Society glossary terms [@Hansell2008] and the CT severity score [@Sonnweber2020; @Luger2022; @Sonnweber2022]. 
Blood biomarkers encompassed hemoglobin and parameters of iron turnover, inflammation and coagulation.

@@

Clustering by the BIPQ items was accomplished by PAM (partitioning around medoids) <span custom-style = 'revision'>with Euclidean distance [@Schubert2019; @Drost2018].</span> 
<span custom-style = 'revision'>The clustering algorithm choice was motivated by its optimal stability in 10-fold cross-validation [@Lange2004] and good explanatory performance measured by the fraction of clustering variance as compared with hierarchical and KMEANS clustering.</span>

@@

The CovILD cohort was recruited <span custom-style = 'revision'>during the first outbreak of SARS-CoV-2</span> between April and June 2020 [@Sonnweber2020]. 
Out of 145 participants enrolled, `r nrow(cov_data$clear_data)` individuals with PSS, lung CT, LFT or cardiological abnormalities, <span custom-style = 'revision'>complete BIPQ and a complete set of study variables set at the one-year follow-up</span> were analyzed. 
The major reasons of patient dropout were missing follow-ups and incomplete BIPQ (__Figure \@ref(fig:fig-consort)__).
<span custom-style = 'revision'>The excluded participants tended towards less severe acute COVID-19, more dermatological and gastrointestinal symptoms at the one-year follow-up, and lower blood hemoglobin and ferritin as compared with the analysis collective; those effects were, however, not significant. 
The excluded participants had also significantly lower SMWD (__Supplementary Figure S3__).</span>

The <span custom-style = 'revision'>analyzed</span> participants were predominantly male (`r get_percent(cov_data$clear_data, "sex")['male']`%), the median age at COVID-19 diagnosis was `r get_median_iqr(cov_data$clear_data, "age")['median']` years (IQR: `r get_median_iqr(cov_data$clear_data, "age")['iqr']`), over one-third were active or ex-smokers (`r get_percent(cov_data$clear_data, "smoking_history")['yes']`%). 
Most participants suffered from comorbidities (`r get_percent(cov_data$clear_data, "comorb_present")['yes']`%), with cardiovascular disease, metabolic and respiratory disorders as leading conditions. 
The participants were classified by COVID-19 severity as ambulatory (`r get_percent(cov_data$clear_data, "cat_WHO")[1]`%), moderate (hospitalized, no intensive care, no oxygen therapy, `r get_percent(cov_data$clear_data, "cat_WHO")[2]`%) and severe COVID-19 survivors (hospitalized, oxygen therapy or intensive care, `r get_percent(cov_data$clear_data, "cat_WHO")[3]`%). 
The ambulatory COVID-19 subset had the lowest median age, smoking and comorbidity rates (__Table \@ref(tab:tab-cohort-baseline)__).

@@

LFT abnormalities affected `r get_percent(cov_data$clear_data, "lufo_red")['yes']`% of participants and were substantially enriched in severe COVID-19 survivors. 
Lung CT abnormalities were found in `r get_percent(cov_data$clear_data, "ct_severity_any")['yes']`% individuals and their frequency and scoring was significantly higher in moderate-to-severe COVID-19 than in the ambulatory subset. 
The leading cardiological finding was low grade diastolic dysfunction (`r get_percent(cov_data$clear_data, "diastolic_dysf")['yes']`%), which was significantly enriched in moderate and severe COVID-19 survivors. 
<span custom-style = 'revision'>Concomitant PSS, abnormal lung CT, LFT and TTE one year after COVID-19 were observed in `r signif(fup_overlap$all_intersection/nrow(cov_data$clear_data), 3) * 100`% of participants. 
In `r signif(14/nrow(cov_data$clear_data), 3) * 100`% of participants, PSS were not accompanied by cardiopulmonary abnormalities (__Supplementary Figure S2__).</span>
Nearly 80% of severe disease survivors attended COVID-19-specific rehabilitation, the rehabilitation rates in the remaining severity strata were below 20% (__Supplementary Table S4__). 
Most laboratory parameters at the one-year follow-up were within their normal values. 
Mild anemia and improper glycemia control (HbA1c) were evident primarily in moderate and severe COVID-19 survivors (__Supplementary Table S5__).

<span custom-style = 'revision'>Females displayed significantly less severe COVID-19 and less chest CT abnormalities but higher PSS rates and counts than males at the one-year follow-up. 
Differences between the genders in frequencies and scoring of specific PSS were not significant (__Supplementary Table S6__ and not shown).</span>

@@

The median <span custom-style = 'revision'>total</span> IP score representing the sum of all BIPQ items was `r get_median_iqr(cov_data$clear_data, "ipq_total")['median']` (IQR: `r get_median_iqr(cov_data$clear_data, "ipq_total")['iqr']`) in the study collective, the differences between ambulatory, moderate and severe COVID-19 were not significant. 
<span custom-style = 'revision'>The emotion/concern/consequences IP component rated by the Q1, Q2, Q5, Q6 and Q8 item sum was the lowest in moderate and the highest in severe COVID-19. 
Moderate COVID-19 convalescents had in turn the highest intensity of the lacking control/coherence IP component scored as the Q3, Q4 and Q7 item sum as compared with the remaining severity subsets; this difference was not significant (__Table \@ref(tab:tab-psych)__).</span>

@@

<span custom-style = 'revision'>The final regularized models explained at least `r (lasso_mod$cv_fit_stats$ipq_total$estimate[4] %>% signif(2) %>% unname) * 100`% and `r (lasso_mod$cv_fit_stats$ipq_total$estimate[12] %>% signif(2) %>% unname) * 100`% of the total IP score variance in the entire data set and cross-validation, respectively. 
Explained variances of the emotion/concern/consequences IP component score were substantially higher with at least `r (eln_mod$cv_fit_stats$ipq_sub1$estimate[4] %>% signif(2) %>% unname) * 100`% in the entire collective and `r (eln_mod$cv_fit_stats$ipq_sub1$estimate[12] %>% signif(2) %>% unname) * 100`% in cross-validation. 
For the lacking control/coherence component, no meaningful models could be established with either of the algorithms (__Supplementary Figure S5__).
The number of explanatory factors with non-zero coefficients in the total IP models varied between `r length(an$variables$ipq_total$lasso_mod)` (LASSO) and `r length(an$variables$ipq_total$eln_mod)` (Elastic Net) (__Figure \@ref(fig:fig-key-factors)A__, __Supplementary Figures S6 - S8__). 
Between `r length(an$variables$ipq_sub1$blass_mod)` (Bayesian LASSO) and `r length(an$variables$ipq_sub1$lasso_mod)` (LASSO) explanatory variables were linked to the emotion/concern/consequences rating (__Figure \@ref(fig:fig-key-factors)B__, __Supplementary Figures S9 - S11__).</span>
<span custom-style = 'revision'>The key variables affecting total IP identified by all three models were PSS count, reduced physical performance and fatigue scoring at the one-year follow-up. 
In uni-variable analysis, each of them was significantly associated with total IP (__Figure \@ref(fig:fig-total-bipq)__). 
The key explanatory factors for the emotion/concern/consequences IP component identified by all three modeling algorithms were 
PSS count, reduced physical performance, fatigue scoring, significant fatigue, hair loss, cough, diastolic dysfunction, elevated C-reactive protein (CRP) at the one-year follow-up as well as rehabilitation during COVID-19 recovery, respiratory comorbidity and body weight class. 
In uni-variable analysis, PSS number, fatigue scoring and significant fatigue, reduced performance, hair loss, rehabilitation and respiratory comorbidity were significantly associated with increased emotion/concern/consequences scores (__Figure \@ref(fig:fig-sub1-bipq)__).</span>

<span custom-style = 'revision'>Age, gender, CT, LFT or TTE abnormalities alone had no significant effects on total IP or its components in uni-variable analysis (__Supplementary Figures S12 - S13__). 
However, lung CT abnormalities at the one-year follow-up amplified the effects of diastolic dysfunction, fatigue and fatigue rating on total IP and the emotion/concern/consequences component, as revealed by significant interactions in two-way ANOVA.
In particular, individuals with concomitant diastolic dysfunction plus CT abnormalities or with high fatigue scoring plus CT abnormalities displayed significantly higher emotion/concern/consequences scores than participants without lung CT findings.
Similar interaction effects of LFT deficits, fatigue and symptom numbers and of diastolic dysfunction and fatigue were discerned as well (__Supplementary Figures S14__ - __S16__).
Participants with respiratory comorbidities had significantly elevated fatigue scores, LFT abnormality rates, total IP and emotion/concern/consequences ratings as compared with the remaining collective. 
This may suggest additional interaction effects of respiratory conditions, LFT deficits and persistent fatigue on IP (__Supplementary Table S7__).</span>

@@

<span custom-style = 'revision'>IP patterns were investigated by unsupervised clustering of the study participants in respect to the BIPQ items. 
Among several algorithms tested, PAM with Euclidean distance [@Drost2018; @Schubert2019] demonstrated the best reproducibility [@Lange2004] and good explanatory performance and was hence used for definition of three IP clusters (__Supplementary Figure S17__).</span>
Roughly half of participants <span custom-style = 'revision'>assigned to the cluster #1 displayed low scores for each BIPQ item.</span> 
This translated to low total IP severity, <span custom-style = 'revision'>, and low scoring of the emotion/concern/consequences and lacking control/coherence IP components.</span>
Another `r signif(ngroups(ipq_clust$clust_obj)$n/sum(ngroups(ipq_clust$clust_obj)$n) * 100, 2)[2]`% of participants assigned to the cluster #2 were characterized by low self-perceived personal or treatment control but also by low rating of the concern, identity, timeline, consequences and emotional representation items. 
<span custom-theme = 'revision'>As a result, the maximum scores of the lacking control/consequence components and low emotion/concern/consequences IP component scores were observed in the cluster #2.</span>
Cluster #3 individuals (`r signif(ngroups(ipq_clust$clust_obj)$n/sum(ngroups(ipq_clust$clust_obj)$n) * 100, 2)[3]`% of participants) had highly elevated rating of the concern, identity, timeline, consequences and emotional representation items as compared with the clusters #1 and #2. 
As a result, the cluster #3 displayed the highest total IP severity, <span custom-style = 'revision'>and the highest emotion/concern/consequence ratings. 
In turn, the scoring  of the lacking control/coherence IP component was low, suggestive of good understanding of the disease and trust in therapy or personal control of COVID-19 sequelae (__Figure \@ref(fig:fig-ipq-clusters)__).
Effect sizes of differences in total IP and its components between the IP clusters were large (Kruskal-Wallis test, total IP: `r clust_chara$test_results$ipq %>% filter(variable == 'ipq_total') %>% .$plot_cap`, emotion/concern/consequences: `r clust_chara$test_results$ipq %>% filter(variable == 'ipq_total') %>% .$plot_cap`, lacking control/coherence: `r clust_chara$test_results$ipq %>% filter(variable == 'ipq_sub2') %>% .$plot_cap`).</span>

<span custom-style = 'revision'>The cluster #1 individuals tended towards the lowest rates of smoking, metabolic and respiratory comorbidities (__Supplementary Table S8__). 
In the cluster #2 the lowest acute COVID-19 severity, rates of rehabilitation and lung CT abnormality grades were observed. 
The cluster #2 tended also towards the lowest frequency of diastolic dysfunction.
By contrast, in the cluster #3 affected by disease-related concerns, the frequency of severe COVID-19, COVID-19-specific rehabilitation rate, chest CT finding severity and PSS frequency at the one-year follow-up were the highest (__Figure \@ref(fig:fig-clust-cov)__). 
In particular, cluster #3 individuals suffered from protracted fatigue, reduced physical performance, dyspnea and sleep problems to the highest extend, and hair loss and gastrointestinal complaints were observed solely in the cluster #3. 
Consequently, the PSS number and fatigue scoring peaked in the cluster #3. 
PSS frequencies and intensity were comparable in the clusters #1 and #2 (__Figure \@ref(fig:fig-clust-symptoms)__). </span>

>>>>

A subset of coronavirus disease 2019 (COVID-19) patients is affected by protracted somatic symptoms, cardiopulmonary pathology and mental health disorders [@Sonnweber2020; @Sahanic2021; @Luger2022; @Sonnweber2022; @Hufner2022; @Al-Aly2021; @Davis2021; @Soriano2022; @Evans2021; @Sudre2021; @Chen2022]. 
Persistent COVID-19-related symptoms have been initially described by patient initiatives [@McCorkell2021] and subsequently recognized as 'post-COVID-19 condition' by the clinical community [@Sonnweber2020; @Sahanic2021; @Davis2021; @Soriano2022; @Al-Aly2021; @Sudre2021]. 
<span custom-style = 'revision'>Most patients recover completely from acute COVID-19 symptoms within four weeks. 
The clearance rate is particularly fast for non-specific infection complaints [@Sudre2021; @Sahanic2021; @Rass2022]. 
In a recent</span> <span custom-style = 'revision2'>meta-analysis</span> <span custom-style = 'revision'>of 1.7 million patients, the pooled prevalence estimate of post-COVID-19 condition, i.e. symptoms present for 3 months or longer, was 43% [@Soriano2022; @Chen2022]. 
Yet, in a rigorously controlled study, approximately 12.7% of SARS-CoV-2-positive individuals suffered from persistent somatic symptoms directly associated with COVID-19 [@Ballering2022].
Fatigue, concentration and memory problems, sleep disorders, physical performance deficits, smell/taste disorders and hair loss, dyspnea and pain are the most common, frequently overlapping and relapsing manifestations [@Chen2022; @Sonnweber2022; @Sahanic2021; @Davis2021; @Rass2022; @Evans2021; @Huang2021; @Ballering2022].</span>

@@

<span custom-style = 'revision'>We aimed to describe IP and IP components, identify demographic and clinical variables affecting IP and explore IP patterns in convalescents suffering from persistent somatic symptoms (PSS), lung or heart abnormalities one year after COVID-19.
To this end, we gauged IP in a long-term observatory cross-sectional COVID-19 collective [@Sonnweber2020; @Sahanic2021; @Luger2022; @Sonnweber2022] with the Brief Illness Perception Questionnaire (BIPQ) [@Broadbent2006] and investigated components, influential factors and patterns of IP by factor analysis, regularized multi-parameter modeling and unsupervised clustering.</span>

@@

<span custom-style = 'revision'>In total, 190 individuals were screened for participation. 
The screened subset size was limited by the number of COVID-19 cases diagnosed at the study centers, their follow-up availability (residency in the study region) and available resources for the follow-up assessments. 
No statistical sample size estimation was performed, as the kinetic of COVID-19 recovery was unknown at the pandemic onset.
Out of 145 enrolled individuals, `r nrow(cov_data$raw_data)` completed the one-year follow-up.</span>
A subset of participants (n = `r nrow(cov_data$clear_data)`) was analyzed who displayed (1) PSS or (2) any abnormality in chest computed tomography or (3) any lung function testing deficits or (4) any heart abnormality in trans-thoracic echocardiography at the one-year follow-up <span custom-style = 'revision'>and had a complete set of BIPQ, demographic and clinical analysis variables (__Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1 - S2__).</span>
<span custom-style = 'revision'>We decided to restrict the analysis to incomplete COVID-19 recovery, because BIPQ was initially designed for and validated in somatic conditions [@Broadbent2006].</span>

@@

Physical performance was rated with the Eastern Cooperative Oncology Group scale. 
Dyspnea was scored with the Modified Medical British Research Council scale.
Fatigue was rated with likert and bimodal Chalder's Fatigue Scales [@Chalder1993; @Morriss1998]. 
Exertional capacity was assessed by six-minute walking distance [@Crapo2012].

The following <span custom-style = 'revision'>PSS</span> were analyzed at the one-year follow-up: 
reduced physical performance (Eastern Cooperative Oncology Group scale $\geq$ 1), 
dyspnea (Modified Medical British Research Council scale $\geq$ 1), 
self-reported cough (yes/no item), 
self-reported sleep problems (yes/no), 
self-reported night sweating (yes/no), 
self-reported hyposmia or anosmia (yes/no), 
self-reported dermatological symptoms (yes/no), 
self-reported gastrointestinal symptoms (yes/no), 
self-reported hair loss (yes/no), 
significant fatigue (bimodal Chalder's Fatigue Scale $\geq$ 4). 
<span custom-style = 'revision'>Records of reduced physical performance, dyspnea, cough, sleep problems, night sweating, hyposmia/anosmia and gastrointestinal symptoms were available for acute COVID-19 and all follow-ups [@Sonnweber2022].</span>

Lung function testing abnormality was defined as at least one parameter < 80% (forced vital capacity, forced expiratory volume in 1 second, total lung capacity, diffusion lung capacity for carbon monoxide) or < 70% (ratio of forced expiratory volume in 1 second to forced vital capacity) of the reference value [@Sonnweber2022]. 
Computed tomography images were evaluated with the Fleischner Society glossary terms [@Hansell2008] and the computed tomography severity score [@Sonnweber2020; @Luger2022; @Sonnweber2022]. 
Blood biomarkers encompassed hemoglobin and parameters of iron turnover, inflammation and coagulation.

@@

Clustering by the BIPQ items was accomplished by the partitioning around medoids algorithm <span custom-style = 'revision'>with Euclidean distance [@Schubert2019; @Drost2018].</span> 
<span custom-style = 'revision'>The clustering algorithm choice was motivated by its optimal stability in 10-fold cross-validation [@Lange2004] and good explanatory performance measured by the fraction of clustering variance as compared with hierarchical and KMEANS clustering.</span>

@@

The CovILD cohort was recruited <span custom-style = 'revision'>during the first outbreak of SARS-CoV-2</span> between April and June 2020 [@Sonnweber2020]. 
Out of 145 participants enrolled, `r nrow(cov_data$clear_data)` individuals with PSS or abnormalities in lung computed tomography, lung function testing or cardiological abnormalities, <span custom-style = 'revision'>complete BIPQ and a complete set of study variables set at the one-year follow-up</span> were analyzed. 
The major reasons of patient dropout were missing follow-ups and incomplete BIPQ (__Figure \@ref(fig:fig-consort)__).
<span custom-style = 'revision'>The excluded participants tended towards less severe acute COVID-19, more dermatological and gastrointestinal symptoms at the one-year follow-up, and lower blood hemoglobin and ferritin as compared with the analysis collective; those effects were, however, not significant. 
The excluded participants had also significantly lower six-minute walking distance (__Supplementary Figure S3__).</span>

The <span custom-style = 'revision'>analyzed</span> participants were predominantly male (`r get_percent(cov_data$clear_data, "sex")['male']`%), the median age at COVID-19 diagnosis was `r get_median_iqr(cov_data$clear_data, "age")['median']` years (interquartile range: `r get_median_iqr(cov_data$clear_data, "age")['iqr']`), over one-third were active or ex-smokers (`r get_percent(cov_data$clear_data, "smoking_history")['yes']`%). 
Most participants suffered from comorbidities (`r get_percent(cov_data$clear_data, "comorb_present")['yes']`%), with cardiovascular disease, metabolic and respiratory disorders as leading conditions. 
The participants were classified by COVID-19 severity as ambulatory (`r get_percent(cov_data$clear_data, "cat_WHO")[1]`%), moderate (hospitalized, no intensive care, no oxygen therapy, `r get_percent(cov_data$clear_data, "cat_WHO")[2]`%) and severe COVID-19 survivors (hospitalized, oxygen therapy or intensive care, `r get_percent(cov_data$clear_data, "cat_WHO")[3]`%). 
The ambulatory COVID-19 subset had the lowest median age, smoking and comorbidity rates (__Table \@ref(tab:tab-cohort-baseline)__).

@@

Lung function testing abnormalities affected `r get_percent(cov_data$clear_data, "lufo_red")['yes']`% of participants and were substantially enriched in severe COVID-19 survivors. 
Abnormalities in lung computed tomography were found in `r get_percent(cov_data$clear_data, "ct_severity_any")['yes']`% individuals and their frequency and scoring was significantly higher in moderate-to-severe COVID-19 than in the ambulatory subset. 
The leading cardiological finding was low grade diastolic dysfunction (`r get_percent(cov_data$clear_data, "diastolic_dysf")['yes']`%), which was significantly enriched in moderate and severe COVID-19 survivors. 
<span custom-style = 'revision'>Concomitant PSS, abnormal lung computed tomography, lung function testing and diastolic dysfunction in echocardiography one year after COVID-19 were observed in `r signif(fup_overlap$all_intersection/nrow(cov_data$clear_data), 3) * 100`% of participants. 
In `r signif(14/nrow(cov_data$clear_data), 3) * 100`% of participants, PSS were not accompanied by cardiopulmonary abnormalities (__Supplementary Figure S2__).</span>
Nearly 80% of severe disease survivors attended COVID-19-specific rehabilitation, the rehabilitation rates in the remaining severity strata were below 20% (__Supplementary Table S4__). 
Most laboratory parameters at the one-year follow-up were within their normal values. 
Mild anemia and improper glycemia control assessed by glycated hemoglobin levels were evident primarily in moderate and severe COVID-19 survivors (__Supplementary Table S5__).

<span custom-style = 'revision'>Females displayed significantly less severe COVID-19 and less chest computed tomography abnormalities but higher PSS rates and counts than males at the one-year follow-up. 
Differences between the genders in frequencies and scoring of specific PSS were not significant (__Supplementary Table S6__ and not shown).</span>

@@

The median <span custom-style = 'revision'>total</span> IP score representing the sum of all BIPQ items was `r get_median_iqr(cov_data$clear_data, "ipq_total")['median']` (interquartile range: `r get_median_iqr(cov_data$clear_data, "ipq_total")['iqr']`) in the study collective, the differences between ambulatory, moderate and severe COVID-19 were not significant. 
<span custom-style = 'revision'>The emotion/concern/consequences IP component rated by the Q1, Q2, Q5, Q6 and Q8 item sum was the lowest in moderate and the highest in severe COVID-19. 
Moderate COVID-19 convalescents had in turn the highest intensity of the lacking control/coherence IP component scored as the Q3, Q4 and Q7 item sum as compared with the remaining severity subsets; this difference was not significant (__Table \@ref(tab:tab-psych)__).</span>

@@

<span custom-style = 'revision'>The final regularized models explained at least `r (lasso_mod$cv_fit_stats$ipq_total$estimate[4] %>% signif(2) %>% unname) * 100`% and `r (lasso_mod$cv_fit_stats$ipq_total$estimate[12] %>% signif(2) %>% unname) * 100`% of the total IP score variance in the entire data set and cross-validation, respectively. 
Explained variances of the emotion/concern/consequences IP component score were substantially higher with at least `r (eln_mod$cv_fit_stats$ipq_sub1$estimate[4] %>% signif(2) %>% unname) * 100`% in the entire collective and `r (eln_mod$cv_fit_stats$ipq_sub1$estimate[12] %>% signif(2) %>% unname) * 100`% in cross-validation. 
For the lacking control/coherence component, no meaningful models could be established with either of the algorithms (__Supplementary Figure S5__).
The number of explanatory factors with non-zero coefficients in the total IP models varied between `r length(an$variables$ipq_total$lasso_mod)` (LASSO) and `r length(an$variables$ipq_total$eln_mod)` (Elastic Net) (__Figure \@ref(fig:fig-key-factors)A__, __Supplementary Figures S6 - S8__). 
Between `r length(an$variables$ipq_sub1$blass_mod)` (Bayesian LASSO) and `r length(an$variables$ipq_sub1$lasso_mod)` (LASSO) explanatory variables were linked to the emotion/concern/consequences rating (__Figure \@ref(fig:fig-key-factors)B__, __Supplementary Figures S9 - S11__).</span>
<span custom-style = 'revision'>The key variables affecting total IP identified by all three models were PSS count, reduced physical performance and fatigue scoring at the one-year follow-up. 
In uni-variable analysis, each of them was significantly associated with total IP (__Figure \@ref(fig:fig-total-bipq)__). 
The key explanatory factors for the emotion/concern/consequences IP component identified by all three modeling algorithms were 
PSS count, reduced physical performance, fatigue scoring, significant fatigue, hair loss, cough, diastolic dysfunction, elevated C-reactive protein at the one-year follow-up as well as rehabilitation during COVID-19 recovery, respiratory comorbidity and body weight class. 
In uni-variable analysis, PSS number, fatigue scoring and significant fatigue, reduced performance, hair loss, rehabilitation and respiratory comorbidity were significantly associated with increased emotion/concern/consequences scores (__Figure \@ref(fig:fig-sub1-bipq)__).</span>

<span custom-style = 'revision'>Age, gender, or abnormalities in chest computed tomography or lung function testing, or diastolic dysfunction alone had no significant effects on total IP or its components in uni-variable analysis (__Supplementary Figures S12 - S13__). 
However, lung abnormalities in computed tomography at the one-year follow-up amplified the effects of diastolic dysfunction, fatigue and fatigue rating on total IP and the emotion/concern/consequences component, as revealed by significant interactions in two-way ANOVA.
In particular, individuals with concomitant diastolic dysfunction plus lung computed tomography abnormalities or with high fatigue scoring plus lung computed tomography abnormalities displayed significantly higher emotion/concern/consequences scores than participants without lung computed tomography findings.
Similar interaction effects of lung function testing deficits, fatigue and symptom numbers and of diastolic dysfunction and fatigue were discerned as well (__Supplementary Figures S14__ - __S16__).
Participants with respiratory comorbidities had significantly elevated fatigue scores, lung function testing abnormality rates, total IP and emotion/concern/consequences ratings as compared with the remaining collective. 
This may suggest additional interaction effects of respiratory conditions, lung function testing deficits and persistent fatigue on IP (__Supplementary Table S7__).</span>

@@

<span custom-style = 'revision'>IP patterns were investigated by unsupervised clustering of the study participants in respect to the BIPQ items. 
Among several algorithms tested, partition aroud medoids with Euclidean distance [@Drost2018; @Schubert2019] demonstrated the best reproducibility [@Lange2004] and good explanatory performance and was hence used for definition of three IP clusters (__Supplementary Figure S17__).</span>
Roughly half of participants <span custom-style = 'revision'>assigned to the cluster #1 displayed low scores for each BIPQ item.</span> 
This translated to low total IP severity, <span custom-style = 'revision'>and low scoring of the emotion/concern/consequences and lacking control/coherence IP components.</span>
Another `r signif(ngroups(ipq_clust$clust_obj)$n/sum(ngroups(ipq_clust$clust_obj)$n) * 100, 2)[2]`% of participants assigned to the cluster #2 were characterized by low self-perceived personal or treatment control but also by low rating of the concern, identity, timeline, consequences and emotional representation items. 
<span custom-theme = 'revision'>As a result, the maximum scores of the lacking control/consequence components and low emotion/concern/consequences IP component scores were observed in the cluster #2.</span>
Cluster #3 individuals (`r signif(ngroups(ipq_clust$clust_obj)$n/sum(ngroups(ipq_clust$clust_obj)$n) * 100, 2)[3]`% of participants) had highly elevated rating of the concern, identity, timeline, consequences and emotional representation items as compared with the clusters #1 and #2. 
As a result, the cluster #3 displayed the highest total IP severity, <span custom-style = 'revision'>and the highest emotion/concern/consequence ratings. 
In turn, the scoring  of the lacking control/coherence IP component was low, suggestive of good understanding of the disease and trust in therapy or personal control of COVID-19 sequelae (__Figure \@ref(fig:fig-ipq-clusters)__).
Effect sizes of differences in total IP and its components between the IP clusters were large (Kruskal-Wallis test, total IP: `r clust_chara$test_results$ipq %>% filter(variable == 'ipq_total') %>% .$plot_cap`, emotion/concern/consequences: `r clust_chara$test_results$ipq %>% filter(variable == 'ipq_total') %>% .$plot_cap`, lacking control/coherence: `r clust_chara$test_results$ipq %>% filter(variable == 'ipq_sub2') %>% .$plot_cap`).</span>

<span custom-style = 'revision'>The cluster #1 individuals tended towards the lowest rates of smoking, metabolic and respiratory comorbidities (__Supplementary Table S8__). 
In the cluster #2 the lowest acute COVID-19 severity, rates of rehabilitation and lung computed tomography abnormality grades were observed. 
The cluster #2 tended also towards the lowest frequency of diastolic dysfunction.
By contrast, in the cluster #3 affected by disease-related concerns, the frequency of severe COVID-19, COVID-19-specific rehabilitation rate, chest computed tomography finding severity and PSS frequency at the one-year follow-up were the highest (__Figure \@ref(fig:fig-clust-cov)__). 
In particular, cluster #3 individuals suffered from protracted fatigue, reduced physical performance, dyspnea and sleep problems to the highest extend, and hair loss and gastrointestinal complaints were observed solely in the cluster #3. 
Consequently, the PSS number and fatigue scoring peaked in the cluster #3. 
PSS frequencies and intensity were comparable in the clusters #1 and #2 (__Figure \@ref(fig:fig-clust-symptoms)__). </span>

<span custom-style = 'revision'>Generally, differences between the IP clusters in PSS readouts like 
PSS number (Kruskal-Wallis test: `r clust_chara$test_results$fup %>% filter(variable == 'sympt_number') %>% .$plot_cap`), 
fatigue rating (`r clust_chara$test_results$fup %>% filter(variable == 'Chalder_FS') %>% .$plot_cap`), 
significant fatigue ($\chi^2$ test: `r clust_chara$test_results$fup %>% filter(variable == 'Chalder_FS-bimodal') %>% .$plot_cap`), 
sleep problems (`r clust_chara$test_results$fup %>% filter(variable == 'sleep_sympt') %>% .$plot_cap`) or 
hair loss (`r clust_chara$test_results$fup %>% filter(variable == 'hair_loss_sympt') %>% .$plot_cap`) and differences in rehabilitation rates (`r clust_chara$test_results$fup %>% filter(variable == 'rehabilitation') %>% .$plot_cap`) were of moderate-to-large effect size.
Effects of differences in 
lung computed tomography abnormality (`r clust_chara$test_results$fup %>% filter(variable == 'ct_severity_any') %>% .$plot_cap`), 
chest computed tomography abnormality scoring (`r clust_chara$test_results$fup %>% filter(variable == 'ct_severity_score') %>% .$plot_cap`), 
lung function testing deficits (`r clust_chara$test_results$fup %>% filter(variable == 'lufo_red') %>% .$plot_cap`) or 
diastolic dysfunction (`r clust_chara$test_results$fup %>% filter(variable == 'diastolic_dysf') %>% .$plot_cap`) were substantially weaker (__Supplementary Table S8__).</span>
